Analysis of SG Flight data
====

## Introduction
We investigate recently released passenger entry count into Singapore.

```{r echo=FALSE}
opts_chunk$set(echo       = TRUE,
               dpi        = 300,
#               fig.width  = 8,
#               fig.height = 5,
               tidy       = TRUE)
#opts_knit$set(root.dir="/export2/home/uesu/reDiamond")
knit_hooks$set(webgl = hook_webgl, htmlcap = function(before, options, envir){
  if(!before) {
    paste('<center><p class="caption">',options$htmlcap,"</p></center>",sep="")
    }
    })
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(rgl)
library(RColorBrewer)
library(preprocessCore)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
library(igraph)
theme_set(theme_bw())
data = read.csv("singaporeFlights.csv")
```

```{r init, echo=FALSE}
#Split date into year and mth 2000M05
data$mth = substr(data$X, start = 6, stop = 7)
data$yr  = substr(data$X, start = 1, stop = 4)

df = data %>%
        select(-c(TOTAL.REGIONS, X))
rownames(df) = data$X
df$date      = data$X

df.long =   df %>%
            gather(area, absCount, -c(date, mth,yr))
```

## Exploratory analyses
We investigate how passenger numbers fluctuate in terms of 

1. Months
2. Months and Years

```{r echo=FALSE, htmlcap="<b>Fig</b>: Boxplots showing human traffic grouped by month", fig.width=8}
qplot(yr, log10(absCount), data = df.long, geom=c("boxplot", "jitter")) +
        labs(x="year", y="Freq (log10)")
```

```{r echo=FALSE,htmlcap="<b>Fig</b>:Same Y scale", fig.width=8}
qplot(date, absCount, data = df.long)      +
        facet_wrap(~area)                  +
        theme(axis.text.x=element_blank()) +
        labs(x="Date", y="Absolute Count")

```{r echo=FALSE, htmlcap="<b>Fig</b>: Free scale", fig.width=8}
qplot(date, absCount, data = df.long)   +
        facet_wrap(~area, scales='free')   +
        theme(axis.text.x=element_blank()) +
        labs(x="Date", y="Absolute Count")
```

```{r echo=FALSE, htmlcap="<b>Fig</b>: Grouped by Year", fig.width=8}
byDate = df.long                  %>%
            group_by(mth,yr,area) %>%
            summarise(total = sum(absCount))

qplot(area, total, data = byDate,geom="bar",stat="identity",fill=mth,position="dodge") +
    facet_wrap(~yr)                                                                    +
    theme(axis.text.x=element_text(angle=90))
```

## Quantile Normalisation
We proceed to normalise the data, by manipulating the data.points to have the same quantiles.

```{r, echo=FALSE}
tonormalise      =  df.long                      %>%
                    select(-date)                %>%
                    spread(yr, absCount, fill=1)
normalised       =  setNames(
                    as.data.frame(normalize.quantiles(as.matrix(tonormalise[,-c(1,2)]))),
                    colnames(tonormalise)[-c(1,2)]
                    )
newDF.long       =  cbind(tonormalise[,c(1,2)], normalised) %>% gather(year, count, -c(mth, area))
newDF.long$date  =  paste0(newDF.long$year, newDF.long$mth)
```

```{r, htmlcap="<b>Fig</b>: Free scale", echo=FALSE}
qplot(year, log10(count), data = newDF.long, geom="boxplot")
```

```{r, htmlcap="<b>Fig</b>: Free scale", echo=FALSE, fig.width=8}
qplot(date, count, data = newDF.long) + facet_wrap(~area)
```

```{r, echo=FALSE}
byDate.norm = newDF.long    %>%
    select(-date)           %>%
    group_by(mth,year,area) %>%
    summarise(total = sum(count))
```

```{r, htmlcap="<b>Fig</b>: Free scale", echo=FALSE, fig.width=8}
#qplot(area, total, data = byDate.norm,geom="bar",stat="identity",fill=mth,position="dodge")+
#facet_wrap(~year)+theme(axis.text.x=element_text(angle=90))
```

```{r}
newDF.wide = newDF.long %>% spread(area, count)
```

## Clustering

We proceed next with clustering the regions,
from which these passengers arrive into groups 
based on how similar they are in terms of their frequencies.

### 1. Hierarchical Clustering

```{r echo=FALSE, }
hc.regions =df                            %>%
    select(SOUTH.EAST.ASIA:OTHER.REGIONS) %>%
    t                                     %>%
    scale                                 %>%
    dist                                  %>%
    hclust
plot(hc.regions, xlab="Regions")
```

### 2. K-means

We plot the elbow plot where we plot the within cluster sum of squares *K*.

```{r echo=FALSE}
wss = rep(0, 10)
for (i in 2:12) 
    wss[i] = sum(kmeans(t(newDF.wide[,-c(1,2,3)]), centers=i, nstart=100, iter.max=1000)$withinss)
```

```{r echo=FALSE, htmlcap="<b>Fig</b>: Elbow plot"}
qplot(x=1:11, y=wss[-1], geom=c("point", "line")) +
ylab("Within groups sum of squares")              +
xlab("Number of centers")
```

From the graph, we see that the best k is 5.

```{r echo=FALSE, message=FALSE}
k               =  kmeans(x=t(df[,2:19]), center= 5, nstart =  25, iter.max=1000)
kmeans.df       =  data.frame(k$cluster)
kmeans.df$name  =  rownames(kmeans.df)
kmeans.df$id    =  1:nrow(kmeans.df)
kmeans.df$col   =  kmeans.df$k.cluster
kmeans.df$col   =  factor(kmeans.df$col, labels= brewer.pal(n=5, name='Set3'))
adjMatrix       =  matrix(rep(0,nrow(kmeans.df)*nrow(kmeans.df)), ncol=nrow(kmeans.df),nrow=nrow(kmeans.df))
unique(kmeans.df$name) %>% sapply(function(namae){
    row = subset(kmeans.df, name == namae)$id
    col = subset(kmeans.df, k.cluster == subset(kmeans.df, name == namae)$k.cluster)$id
    adjMatrix[row,col]<<-1
    })
```

Presenting each cluster as a group in the network.

```{r, echo=FALSE}
plot(simplify(
              graph.adjacency(adjmatrix=adjMatrix)),
     vertex.label=kmeans.df$name, vertex.color=kmeans.df$col)
```

### Singular Vector Decomposition

Finally, we apply SVD to the the data and look for the overarching patterns.

```{r echo=FALSE, htmlcap="<b>Fig</b>: PC1"}
svdObj = svd(scale(df %>% select(SOUTH.EAST.ASIA:OTHER.REGIONS)))
grid.arrange(
             qplot(1:length(svdObj$u[,1]), svdObj$u[,1], geom=c("point","line"))+labs(title="PC1"),
             qplot(1:length(svdObj$u[,2]), svdObj$u[,2], geom=c("point","line"))+labs(title="PC2"),
             qplot(1:length(svdObj$u[,3]), svdObj$u[,3], geom=c("point","line"))+labs(title="PC3")
             )
clus = kmeans(svdObj$v[,1:3], centers=3, nstart=25, iter.max=1000)$cluster
```

```{r webgl=TRUE, eval=TRUE, echo=FALSE}
plot3d(svdObj$v[,1], svdObj$v[,2], svdObj$v[,3], size=10, col=clus)
text3d(svdObj$v[,1:3], text=colnames(df)[1:19], cex = 1,col=clus)
```

```{r echo=FALSE, htmlcap="<b>Fig:</b>: Applying Hierarchical Clustering on PCS"}
rownames(svdObj$v) <- colnames(df)[1:19]
pca.hc = hclust(dist(svdObj$v[,1:3]))
plot(pca.hc)
```
